---
title: 'Air Crash Investigation: How does Flight Conditions affect Fatality?'
author: "Erika Lu & Sophie Wu"
date: '2020-12-14'
output:
  github_document:
    toc: true
---

## Purpose

In this project, we focus on studying the air crash investigation dataset from National Transportation Safety Board (NTSB). We want to study the question: How well can flight parameters (engine type/number, weather condition, the broad phase of flight, etc.) be used to predict the fatality of a flight? 


## Background 

The dataset we used contained information on air crashes investigated by the NTSB from 1982 to 2007. The investigation range of NTSB includes all air crashes happened within U.S. territory, or when the involved aircraft are U.S. made, or the aircraft carries passengers/crew members that are U.S. citizens, or otherwise invited by the transportation safety bureau from another country. 

The dataset includes information on not only military and commercial flights but general aviation as well. It is easier to get a piloting license for general aviation than other types of aviation and it is generally regarded as more dangerous because of lower safety standards.

## Uncertainties

The data on fatalities are collected from air crash investigation reports, and we may be able to assume that there are no sampling uncertainties in the collection of data, because our dataset is pretty close to the population. However, we also believe that there's a underestimated of the number of air crashes: as we look through the dataset, it seems that not all of the crashes are represented in the dataset. For example, during the terrorist attack on September 11, 2001, four airplanes are hijacked and crashed, and the dataset only records one of them. We do not know how many other accidents went unreported.

## Setup

```{r}
library(tidyverse)
library(modelr)
library(broom)
library(scales)
library(gganimate)
```

## Loading the dataset

```{r}
filename  <- "./data/AviationData.txt"

df_aviation <- read.delim2(filename, header = TRUE, sep = "|", dec = ".")

df_aviation <- df_aviation %>%
  filter(str_detect(Aircraft.Category, "Airplane"), str_detect(Investigation.Type, "Accident")) %>%
  mutate(
    IsGeneralAviation = str_detect(FAR.Description, "91")
  )


```

## Initial inspection of the dataset

```{r}
df_aviation %>% glimpse()
```

**Observations:**

A glimpse at the dataset shows that 6,486 accidents and incidents are recorded in this dataset. 32 columns are presented (+1 added), in which the total number of fatal, serious, and minor injuries, as well as the total uninjured number are recorded as output variables, and other variables including `Make`, `Model`, `Number.of.Engines`, `Engine.Type`, `Purpose.of.Flight` and `Broad.Phase.of.Flight` are interpreted as input variables in this study. 

```{r}
df_binomial <- df_aviation %>% mutate(Fatal = as.logical(Total.Fatal.Injuries))
df_binomial %>% count(Fatal)
```

**Observations:**

A simple check on the `Total.Fatal.Injuries` column of the dataset reveals that from 1982 to 2007, there are 2505 non-fatal crashes and 943 fatal crashes, and 3038 flight crash information are not available. 

```{r}
df_aviation %>% count(IsGeneralAviation)
```
**Observations:**
Most of the recorded accidents, 5612 of them, are general aviation flights. Other flights make up the rest of the 874 of them.

## Visualizations

```{r}
#install.packages("randomcoloR")

library(randomcoloR)

# Styling
blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

df_aviation %>%
  filter(Broad.Phase.of.Flight != "  ") %>%
  group_by(Broad.Phase.of.Flight, IsGeneralAviation) %>%
  summarize(
    count = n(),
  ) %>%
  ggplot() +
  geom_bar(mapping = aes(x = "", y = count, fill = Broad.Phase.of.Flight ), stat = "identity", position = position_fill()) +
  
  # Styling 
  blank_theme +
  theme(
    axis.text.x=element_blank()
        ) +
    labs(x = "Month",
        y = "Accidents",
        fill = "Flight Type") +
  scale_fill_manual(values=distinctColorPalette(12))+
  coord_polar(theta = "y", start = 0) +
  
  #Facet
  facet_wrap(~ ifelse(IsGeneralAviation, "General Aviation (Total: 5581)", "Other (Total: 660)"))
```

**Observations:**

Accidents happen more often during certain phases when General Aviation is compared to other types of flights. A dangerous phase for GA is Landing, while not many accidents happen in that phase for other types of flights. This may be due to different dynamics when it comes to smaller planes. GA pilots may also fly in a more risky manner than commercial pilots. GA may be less accident prone during maneuvering because the smaller planes were built to do such maneuvers other planes like passenger planes/cargo planes were are not suited to do such maneuvers and when they do it is because of an emergency.


### Air crashes and dates


```{r}
df_aviation_dates <-df_aviation %>%
  mutate(
    Month.Day = substr(Event.Date,2,6),
    Year = substr(Event.Date, 8, 11),
    Date = paste(Year, substr(Event.Date, 2, 3), substr(Event.Date,5,6), sep = "-", collapse = NULL),
    Month = paste("2020", substr(Event.Date, 2, 3), "01", sep = "-", collapse = NULL)
  ) 

df_aviation_dates$Month.Day = as.Date(df_aviation_dates$Month.Day, format = "%m/%d")
df_aviation_dates$Date = as.Date(df_aviation_dates$Date, format = "%Y-%m-%d")
df_aviation_dates$Month = as.Date(df_aviation_dates$Month, format = "%Y-%m-%d")

```

```{r}
df_aviation_dates%>%
  ggplot() +
  geom_boxplot(mapping = aes(x = factor(format(Month, "%m")), y = Total.Fatal.Injuries, color = ifelse(IsGeneralAviation, "General Aviation", "Commercial + Other")) ) +
  labs(x = "Month",
      y = "Deaths",
      color = "Flight Type") +
  ggtitle("Fatalities Per Accident") +
  scale_y_log10()
```
**Observations:**
Commercial + Other flights tend to have higher fatalities every month compared to General Aviation. This may be due to the fact that GA planes tend to be much smaller and usually carries the pilot and a handful of passengers. It is difficult to see a trend of how month affects the fatality of an accident but it looks like there may be slightly more fatalies per accident in the months of August and October for Commercial + Other flights. 


```{r}
df_aviation_dates %>%
  count(Month.Day) %>%
  ggplot() +
  geom_point(mapping = aes(x = Month.Day, y = n )) +
  scale_x_date(date_breaks = "1 month", date_minor_breaks = "1 week",
             date_labels = "%B") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggtitle("Accidents by Time of Year")
```
**Observations:**

```{r}
# AGGREGATED BY MONTH

#preprocess data
df_month <- df_aviation_dates %>%
  group_by(Month, Year, IsGeneralAviation) %>%
  mutate(
    count = n(),
    average = mean(n())
  ) %>%
  ungroup() %>%
  arrange(Date) 

df_month$frame <- seq.int(nrow(df_month))

fig <- df_month %>%
  ggplot(aes(color = ifelse(IsGeneralAviation, "General Aviation", "Commercial + Other"))) +
  theme_bw() +
  labs(title = "Average Monthly Plane Accidents") +
  geom_line(data = df_month %>% filter(IsGeneralAviation), aes(x = Month, y = average, group = Year), size = 0.15) +
  geom_line(data = df_month %>% filter(!IsGeneralAviation), aes(x = Month, y = average, group = Year), size = 0.15) +
   labs(x = "Month",
        y = "Accidents",
        color = "Flight Type") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b", expand = expansion(add = c(0, -15))) +
  geom_text(aes(x = as.Date("2020-10-01"), y = 260, label = Year), position = position_nudge(y = -0.1, x = 0), size = 10, color = "gray20") +
  transition_reveal(along = frame, keep_last = FALSE)


# animate(fig, nframes = 200, fps = 10, width=1200,height=800, res = 130,
#         render = av_renderer("Monthly Airplane Accidents.mp4", codec = "libx264"))

animate(fig, renderer = gifski_renderer(), width=600, height=400)

```


**Observations:**

The last couple of graphs look at occurences of accidents rather than fatality to better understand the dataset. 

One reason we may see more accidents in the middle of the year is that more people have time to fly (whether for hobby or travel) during the summer months due to the United States vacation schedule. More flights lead to more accidents. If the scope of our project was wider, we would analyze data on all flights, rather than just the accidents. 

Another reason that may explain why accidents occur more often in June and July: during that time thunderstorms are more frequent and more severe, and microbursts can occur, leading to poor visibility and unstable flights (especially at landing).

### Geography


```{r}
library("ggplot2")
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")

world <- ne_countries(scale = "medium", returnclass = "sf")

blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold"),
  )

df_geo <- df_aviation_dates %>%
  select (
    Year,
    Longitude,
    Latitude,
    IsGeneralAviation
  ) %>%
  drop_na()


ggplot(data = world) +
  geom_sf() +
  geom_point(data = df_geo, aes(x = Longitude, y = Latitude, fill = ifelse(IsGeneralAviation, "General Aviation", "Commercial + Other")), size = 1, 
      shape = 23) +
  coord_sf(xlim = c(-130, -65), ylim = c(20, 60), expand = FALSE) +
  blank_theme +
  theme(
    axis.text.x=element_blank(),
    axis.text.y=element_blank()
    ) +
  labs(fill = "Flight Type") +
  ggtitle("Aviation Accident Sites")
```
**Observations:**
While there is no discernible trend for Commercial + Other type accidents, we see that General Aviation accidents happen most often along the coasts and in the north east.

This may be due to confounding variables such as population and wealth that would increase the prevalence of GA pilots on the coast. Or perhaps planes are more prone to accidents in coastal weather. 

### Weather conditions

```{r}
df_binomial %>%
  filter(is.na(Fatal) == FALSE) %>% 
  group_by(Weather.Condition) %>% 
  count(Fatal) %>%
  ggplot(aes(x = Weather.Condition, y = n, fill = Fatal)) +
  geom_col(position = "dodge") +
  ggtitle("Number of Fatal and Non-fatal Crashes in Different Weather Conditions")
```
```{r}


df_binomial %>%
  filter(!is.na(Fatal), IsGeneralAviation) %>%
  group_by(Fatal, IsGeneralAviation, Weather.Condition) %>%
  summarize(
    count = n(),
  ) %>%
  ggplot() +
  geom_bar(mapping = aes(x = "", y = count, fill = Fatal ), stat = "identity", position = position_fill()) +
  
  # Styling 
  blank_theme +
  theme(
    axis.text.x=element_blank()
        ) +
    labs(fill = "Fatal") +
  coord_polar(theta = "y", start = 0) +
  #Facet
  facet_wrap(~Weather.Condition) +
  ggtitle("Flight Fatalities (General Aviation)")

df_binomial %>%
  filter(!is.na(Fatal), !IsGeneralAviation) %>%
  group_by(Fatal, IsGeneralAviation, Weather.Condition) %>%
  summarize(
    count = n(),
  ) %>%
  ggplot() +
  geom_bar(mapping = aes(x = "", y = count, fill = Fatal ), stat = "identity", position = position_fill()) +
  
  # Styling 
  blank_theme +
  theme(
    axis.text.x=element_blank()
        ) +
    labs(fill = "Fatal") +
  coord_polar(theta = "y", start = 0) +
  #Facet
  facet_wrap(~Weather.Condition) +
  ggtitle("Flight Fatalities (Non General Aviation)")

```


**Observations:**

VMC, or visual meteorological conditions, means that pilots have sufficient visibility to fly the aircraft maintaining visual separation from terrain and other aircraft. IMC, or instrumental meteorological conditions, means that pilots need to fly primarily by reference to instruments, and therefore under instrument flight rules (IFR). Typically, IMC means flying in cloudy or bad weather, while VMC means fine weather. 

Comparing the absolute value of fatal crashes in VMC is higher than IMC. However, this may be due to the fact that much more flights are flown in VMC, as flights are likely to be canceled or diverted in deteriorating weather. Comparing the ratio of fatal and non-fatal crashes in VMC and IMC, it turns out that in IMC the crashes have a higher odds of being fatal. 

The contribution of weather condition to a fatal crash will be investigated in later sections by fitting it into a logistic model. 

### Aircraft damage

```{r}
df_binomial %>%
  filter(is.na(Fatal) == FALSE) %>% 
  group_by(Aircraft.Damage) %>% 
  count(Fatal) %>%
  ggplot(aes(x = Aircraft.Damage, y = n, fill = Fatal)) +
  geom_col(position = "dodge") +
  ggtitle("Number of Fatal and Non-fatal Crashes for different Damage")
```

**Observations:** 

We can reasonably assume that, when an aircraft is severely damaged or destroyed, there is a higher probability of having fatalities, while for minor and substantial damages the fatality is much lower. This figure above proves that, the destroyed air crashes are much more fatal comparing to other categories. 

The `Aircraft.Damage` criteria will be studied in later sections where a model is fitted to figure out how much the aircraft damage contributes to fatality. 


## Modeling

### A model calculating the effect of weather condition on fatality

```{r}
fit_weather <-
  glm(
    formula = Fatal ~ Weather.Condition,
    data = df_binomial %>%
      filter(
        !is.na(Fatal), Weather.Condition %in% c(" IMC ", " VMC ", " UNK ")
      ) %>%
      mutate(Weather.Condition = fct_relevel(Weather.Condition, " UNK ")),
    family = "binomial"
  )

fit_weather %>% tidy()
```

**Observations:**

According to the model, setting `UNK` as a reference, `VMC` has a much lower estimate comparing to `IMC`. This result is consistent with our conclusion from the bar chart: in IMC conditions, there is a higher odds of the air crash being fatal. The negative estimate for `VMC` 

### A model calculating the predicting effect of aircraft damage

```{r}
 fit_damage <-
  glm(
    formula = Fatal ~ Aircraft.Damage,
    data = df_binomial %>%
      filter(
        !is.na(Fatal), Aircraft.Damage %in% c(" Destroyed ", " Minor ", " Substantial ")
      ) %>%
      mutate(Aircraft.Damage = fct_relevel(Aircraft.Damage, " Substantial ")),
    family = "binomial"
  )

fit_damage %>% tidy()
```

**Observations:**

The estimate for destroyed aircraft for modeling fatality is much higher, while there's no significant difference between minor damages and substantial damages (0.13 vs 0). This confirms our hypothesis of when an aircraft is severely damaged or destroyed, there is a higher probability of having fatalities. A destroyed aircraft is strongly correlated with fatalities. 



### Fitting a model with two variables

We noticed from the previous figure that in June, July and August, there is a peak in air crashes. We thus made a hypothesis that the more frequent severe weather including thunderstorms are microbursts during that time period contributes to the higher frequency of air crashes. 

```{r}

df_date<-df_aviation %>%
  filter(str_detect(Aircraft.Category, "Airplane"), str_detect(Investigation.Type, "Accident")) %>%
  mutate(
    Month.Day = substr(Event.Date,2,6),
    Year = substr(Event.Date, 8, 11),
    Date = paste(Year, substr(Event.Date, 2, 3), substr(Event.Date,5,6), sep = "-", collapse = NULL),
    Month = substr(Event.Date, 2, 3)
  ) %>%
  mutate(Month = as.integer(Month)) %>%
  mutate(Fatal = as.logical(Total.Fatal.Injuries))
  
df_date %>% summary()
  fit_date <-
   glm(
     formula = Fatal ~ Month + Weather.Condition,
     data = df_date %>%
       filter(
         !is.na(Fatal), 
         Month %in% c(6, 7, 8),
         Weather.Condition %in% c(" IMC ", " VMC ", " UNK ")
       ) %>%
       mutate(Weather.Condition = fct_relevel(Weather.Condition, " UNK ")),
     family = "binomial"
   )
 
 fit_date %>% tidy()
```

**Observations:**

When the time of crashes `Month` is controlled, the estimate for `IMC` decreases and becomes negative, indicating that `IMC` is less effective at predicting fatality during summer. We can thus reject our hypothesis that severe weather is positively correlated to a higher fatality. 

## Summary

In this report, we studied air crashes investigated by NTSB from 1982 to 2007, and we identified variables that can potentially be used for predicting air crash fatalities. We use weather and aircraft damage as the main input variables to predict if a plane crash is fatal, and we figure out the correlation between these variables and a fatal crash. 

For future studies, we would like to include more variables (e.g. the cause of crashes) that are not available in this dataset.





